---
title: "ADTA 5410 RLab1"
author: "Pavan Kumar Bodepudi"
date: "04/18/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
You are expected to complete this homework in R by using this Rmarkdown file in RStudio environment.   Create a new folder in your computer and save this Rmarkdown file and the Data_RLab1.csv file in the same folder. 


Submission instructions: Submit your knitted Rmarkdown file along with your answers in Canvas.   


0. Write your full name in the "Author" part above and save as this Rmarkdown file by adding your initials to the end (i.e., RLab1_LB.Rmd) 



## Intro 

In this Lab assignment, we will use data from the U.S. Dept. of Education College scorecard. Data is in csv file. We will import it to R format and name the dataset as **school_raw** by using the **read.csv** command, shown below, and in cases where any cell takes the value of "NULL" or "PrivacySuppressed", we will code them NA. 


```{r }
if (!require("knitr")) install.packages("knitr")
if (!require("dplyr")) install.packages("dplyr")
library(knitr)
library(dplyr)
school_raw <- read.csv("Data_RLab1.csv", na = c("NULL", "PrivacySuppressed"))

```

Data is messy. There are `r dim(school_raw)[1]` observations and `r dim(school_raw)[2]` variables.   


```{r }

dim(school_raw)
```

For the sake of this lab assignment, we will only keep the following 6 variables.  


- CONTROL - onwnership type (public, private - NP, private - P)
- C150_4 - Completion/graduation  rate
- PREDDEG - Predominat type of degree awarded
- SAT_AVG - average SAT score of students admitted
- UGDS - Enrollment of undergraduat certificate/degree-seeking students
- COSTT4_A - Average cost of attendance (academic year institutions)


We will use the following code to keep only the above 6 variables and call the reduced dataset *schools*. 

```{r }

schools  <- subset(school_raw, select =c(CONTROL,  C150_4, PREDDEG,SAT_AVG, UGDS, COSTT4_A))

head(schools)
```





# Prepare the Data 
1. **[Prepare the Data #1]**  Filter *schools* data by  keeping only predominantly bachelor's degree granting schools ($PREDDEG=='3'$) and call it *schools_filtered*. **What are the dimensions of the filtered data *schools_filtered* (number of rows and columns)**
 


```{r }
# Your code goes in here
schools_filtered<-schools[schools$PREDDEG=='3',]
dim(schools_filtered)
```
2.    **[Prepare the Data #2]** How many missing values are there for each variable in the filtered data set *schools_filtered*? Write an R code below to calculate the total number of missing values for each variable. 
```{r }
# Your code goes in here
summary(schools_filtered)
colSums(is.na(schools_filtered))
sapply(schools_filtered, function(x) sum(is.na(x)))
```

3.  **[Prepare the Data #3]** Now, write an R code to drop observations in *schools_filtered* when C150_4 variable has missing values and call it *schools_filtered1* .  What are the dimensions of *schools_filtered1* data (number of rows and columns)?    

```{r }
# Your code goes in here
schools_filtered1<-subset(schools_filtered, !is.na(schools_filtered$C150_4)) 
dim(schools_filtered1)

```


4.  **[Prepare the Data #4]** Now, impute (replace) missing values in **schools_filtered1** **SAT_AVG** and **COSTT4_A** with their corresponding median scores. When you calculate the median for each variable, ignore the NAs. What are the mean values of **SAT_AVG** and **COSTT4_A**  after imputation?


```{r }
# Your code goes in here
median_SAT_AVG<-median(schools_filtered1$SAT, na.rm = TRUE)
median_COSTT4_A<-median(schools_filtered1$COSTT4_A, na.rm = TRUE)
mean_SAT_AVG<-mean(schools_filtered1$SAT, na.rm = TRUE)
mean_COSTT4_A<-mean(schools_filtered1$COSTT4_A, na.rm = TRUE)
median_SAT_AVG
median_COSTT4_A
mean_SAT_AVG
mean_COSTT4_A
library(knitr)
schools_filtered1$SAT[is.na(schools_filtered1$SAT)]<-median_SAT_AVG
schools_filtered1$COSTT4_A[is.na(schools_filtered1$COSTT4_A)]<-median_COSTT4_A
schools_filtered1$SAT[is.na(schools_filtered1$SAT)]<-mean_SAT_AVG
schools_filtered1$COSTT4_A[is.na(schools_filtered1$COSTT4_A)]<-mean_COSTT4_A
#summary(schools_filtered1)


```


5. **[Prepare the Data #5]**  Now, drop ***PREDDEG**  from the data set and name your dataset as **schools_filtered2**. 
Also,  rename  variables in **schools_filtered2** as follows:   
Rename SAT_AVG  to SAT

Rename C150_4   to graduation_rate

Rename UGDS  to enrollment.

Rename COSTT4_A  to cost.

Then, write an R code to treat CONTROL variable as a factor.  After that, run the following code in your console and choose from below the corresponding code output:
**schools_filtered2$CONTROL<-as.factor(schools_filtered2$CONTROL)**

```{r }
# Your code goes in here

#cor(schools_filtered2$CONTROL, schools_filtered2$cost)
# Your code goes in here
colnames(schools_filtered1)[which(names(schools_filtered1) == "SAT_AVG")] <- "SAT"
colnames(schools_filtered1)[which(names(schools_filtered1) == "C150_4")] <- "graduation_rate"
colnames(schools_filtered1)[which(names(schools_filtered1) == "UGDS")] <- "enrollment"
colnames(schools_filtered1)[which(names(schools_filtered1) == "COSTT4_A")] <- "cost"
schools_filtered2 <- subset(schools_filtered1, select =c(CONTROL, graduation_rate, SAT,enrollment,cost))
dim(schools_filtered2)
schools_filtered2$CONTROL <- as.factor(schools_filtered2$CONTROL)
str(schools_filtered2)

```


6. **[Prepare the Data #6]** 
For the **enrollment** variable in **schools_filtered2**, calculate the followings:
- The first quartile (Q1) for enrollment
- The third quartile (Q3) for enrollment
- Number of observations in enrollment data that can be considered extremely high (that are larger than 3rd quartile for enrollment+1.5*Interquartile range for enrollment)

```{r }
# Your code goes in here
quantile(schools_filtered2$enrollment,prob=c(.25,.5,.75))
sum(schools_filtered2$enrollment>(4951+(1.5)*4043))

```



Make sure that your R codes will run. Knit your file to html and submit it on Canvas along with your answers.  